---
title: "Simulation-based inference - hypothesis testing"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>", highlight = TRUE,
                      fig.align = "center")
```

## Main ideas

- Understand the hypothesis testing framework

- Understand the statistical conclusions that can and cannot be made from a
  hypothesis test
  
- Use functions from `infer` to carry-out a simulation-based hypothesis test
  for the population mean and population proportion
  
# Packages

```{r packages}
library(tidyverse)
library(infer)
```

# Data

We'll continue to work with the sample of Zoom screen-time data we obtained.
To make things easier with the `infer` functions, we'll create a tibble with
`time` as a single variable.

```{r zoom_data}
zoom <- tibble(
  time = c(299, 192, 196, 218, 194, 250, 183, 218, 207, 
           209, 191, 189, 244, 233, 208, 216, 178, 209, 
           201, 173, 186, 209, 188, 231, 195, 200, 190, 
           199, 226, 238)
)
```

```{r preview_zoom}
zoom
```

# Set seed

To obtain reproducible results, set the seed for the random number generation.

```{r set_seed}
set.seed(1421)
```

# Notes

Recall our hypothesis testing framework:

1. Start with two hypotheses about the population: the null hypothesis and the 
   alternative hypothesis.

2. Choose a (representative) sample, collect data, and analyze the data.

3. Figure out how likely it is to see data like what we observed, **IF** the 
   null hypothesis were in fact true.

4. If our data would have been extremely unlikely if the null claim were true, 
   then we reject it and deem the alternative claim worthy of further study. 
   Otherwise, we cannot reject the null claim.
   
## Example: testing population mean - $\mu$
   
We've already done items 1 and 2, where

$$H_0: \mu = 200$$
$$H_1: \mu \neq 200$$

For this study, let $\alpha = 0.05$. 

To tackle items 3 and 4, we'll use a simulation-based approach with functions
from `infer`.

### Simulate the null distribution

Recall that there is variability in the sampling distribution of the sample
mean. We need to account for this in our statistical study. Just as we did
for confidence intervals, we'll use a bootstrap procedure here.

1. `specify()` the variable of interest

2. set the null hypothesis with `hypothesize()`

3. `generate()` the bootstrap samples

4. `calculate()` the statistic of interest

```{r zoom_null}
null_dist <- zoom %>% 
  specify(response = time) %>% 
  hypothesize(null = "point", mu = 200) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
  calculate(stat = "mean")
```

### Visualize the null distribution

```{r zoom_null_viz}
visualise(null_dist) +
  labs(x = "Sample means", y = "Count", title = "Simulated null distribution")
```

**What do you notice?**

### Compute p-value

Next, we calculate the probability of getting data like ours, *or more extreme*,
if $H_0$ were in fact actually true.

Our observed sample mean is 209 minutes.

```{r zoom_xbar}
x_bar <- zoom %>% 
  summarise(mean_time = mean(time))

x_bar
```

```{r zoom_null_viz_pvalue}
visualise(null_dist) +
  shade_p_value(obs_stat = x_bar, direction = "two-sided") +
  labs(x = "Sample mean", y = "Count")
```

In the context of this simulation-based approach, the p-value is the proportion
of observations shaded light-red. To compute this, `infer` provides a 
convenient function -- `get_p_value()`.

```{r zoom_pvalue}
null_dist %>% 
  get_p_value(obs_stat = x_bar, direction = "two-sided")
```

### Conclusion

Given the calculated p-value and the specified $\alpha$, what conclusion do 
you make?

## Practice 1

Recall our original example: The state of North Carolina claims that students in
8th grade are spending, on average, 200 minutes on Zoom each day. Suppose in
reporting this the incorrect metric was specified, it should have been the
median time. Use your sample data to investigate if the median Zoom screen-time
is more than 200 minutes.

(1) Write out the hypotheses for this statistical test. Let $M$ represent the
    population median. Let $\alpha = 0.05$.
    
(2) Generate the null distribution.

```{r practice_1_2}
null_dist_med <- zoom %>% 
  specify(response = time) %>% 
  hypothesize(null = "point", med = 200) %>% 
  generate(reps = 10000) %>% 
  calculate(stat = "median")
```

(3) Visualize the null distribution, observed statistic, and shaded region
    corresponding to the p-value.
    
```{r practice_1_3}
med <- zoom %>% summarise(med_time = median(time))

visualise(null_dist_med) +
  shade_p_value(obs_stat = med, direction = "greater") +
  labs(x = "Medians", y = "Count")
```

(4) Interpret the results of your test in the context of the data.

```{r practice_1_4}
null_dist_med %>% 
  get_p_value(obs_stat = med, direction = "greater")
```

## Example: testing population proportion - $p$

People providing an organ for donation sometimes seek the help of a special 
medical consultant. These consultants assist the patient in all aspects of the
surgery, with the goal of reducing the possibility of complications during the 
medical procedure and recovery. Patients might choose a consultant based in 
part on the historical complication rate of the consultant's clients.

One consultant tried to attract patients by noting that the average complication 
rate for liver donor surgeries in the US is about 10%, but her clients have 
only had 3 complications in the 62 liver donor surgeries she has facilitated. 
She claims this is strong evidence that her work meaningfully contributes to 
reducing complications (and therefore she should be hired!).

(1) Write out the hypotheses for this statistical test. Let $p$ represent the
    population proportion of complications from liver donor surgeries. State
    your significance level.
    
    $$H_0: p = 0.10$$
    $$H_A: p < 0.10$$
    
    Let $\alpha = 0.01$.
    
(2) Generate the null distribution.

```{r liver_data}
liver <- tibble(
  surgery_result = rep(c("complication", "no complication"), times = c(3, 59))
)

liver
```

```{r liver_null}
null_dist_phat <- liver %>% 
  specify(response = surgery_result, success = "complication") %>% 
  hypothesise(null = "point", p = 0.10) %>% 
  generate(reps = 1000) %>% 
  calculate(stat = "prop")
```

(3) Visualize the null distribution, observed statistic, and shaded region
    corresponding to the p-value.
    
```{r liver_null_viz}
p_hat <- liver %>% 
  count(surgery_result) %>% 
  mutate(prop = n / sum(n)) %>% 
  filter(surgery_result == "complication") %>% 
  select(prop)

visualise(null_dist_phat) +
  shade_p_value(obs_stat = p_hat, direction = "less") +
  labs(x = "Sample proportion", y = "Count")
```

(4) Interpret the results of your test in the context of the data.

```{r liver_pvalue}
null_dist_phat %>% 
  get_p_value(obs_stat = p_hat, direction = "less")
```

## Practice 2

Consider the mice data from the previous notes.

```{r data_mice}
mice <- read_table("http://users.stat.ufl.edu/~winner/data/micerad.dat",
                   col_names = FALSE) %>% 
  rename(dose = X1, treat = X2, died = X3)
```

Previous studies have shown that 50% of mice die when subject to radiation
despite being on a treatment. Does the Streptomycin Therapy 
(treatment in this study) produce a survival rate better than 50%? Perform
a statistical hypothesis test to investigate. State the hypotheses,
significance level, p-value, and conclusion. What is a hidden variable we
are not considering when conducting this test?

$$H_0: p = 0.50$$
$$H_A: p < 0.50$$

Let $\alpha = 0.01$.

First, we'll compute the observed sample proportion of mice that died while
on the treatment.

```{r mice_p_hat}
p_hat <- mice %>% 
  filter(treat == 1) %>% 
  summarise(mean_prop = mean(died))

p_hat
```

```{r mice_test}
mice %>% 
  filter(treat == 1) %>% 
  mutate(outcome = ifelse(died == 1, "died", "survived")) %>% 
  specify(response = outcome, success = "died") %>% 
  hypothesise(null = "point", p = 0.50) %>% 
  generate(reps = 10000) %>% 
  calculate(stat = "prop") %>% 
  get_p_value(obs_stat = p_hat, direction = "less")
```

Given that the p-value is less than $\alpha$ we reject the null hypothesis.
That is, we reject the claim that the survival rate of mice exposed to radiation
while on the Streptomycin Therapy is 50%.

A hidden variable we haven't accounted for and assumed to be constant was the
radiation `dose`. In our example, this is okay because our main focus is 
understanding the testing framework. However, in practice, it would be bad
to not consider this in an analysis.

## References

1. C.W. Hammond, et al. (1955). "The Effect of Streptomycin Therapy
   on Mice Irradiated with Fast Neutrons", Radiation Research, Vol2,#4,
   pp.354-360

2. "Infer - Tidy Statistical Inference". Infer.Netlify.App, 2021, 
   https://infer.netlify.app/index.html.
   
